A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-06-11 inst/manuscript/manuscript.tex
added zenodo reference

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less (be
more precise) than parameter estimates from samples lacking spatial
balance (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm to see widespread use was the generalized random tessellation
stratified (GRTS) algorithm (Stevens and Olsen, 2004). After the GRTS
algorithm was developed, several other spatially balanced sampling
algorithms emerged, including stratified sampling with compact
geographical strata (Walvoort et al., 2010), the local pivotal method
(Grafström et al., 2012; Grafström and Matei, 2018), spatially
correlated Poisson sampling (Grafström, 2012), balanced acceptance
sampling (Robertson et al., 2013), within-sample-distance sampling
(Benedetti and Piersimoni, 2017), and Halton iterative partitioning
sampling (Robertson et al., 2018). In this manuscript, we select
spatially balanced samples using the GRTS algorithm because it is
readily available in the spsurvey \textbf{\textsf{R}} package (Dumelle
et al., 2022) and naturally accommodates finite and infinite sampling
frames, unequal inclusion probabilities, and replacement units.
Replacement units are additional population units that can be sampled
when a population unit originally selected can no longer be sampled. A
couple of reasons why an originally selected site can no longer be
sampled include its location being physically inaccessible or it is on
private land that the researcher does not have permission to access.

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate variances. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the response value of the \(i\)th
sampled population unit, \(\bar{z}\) is the mean of all \(z_i\), \(n\)
is the sample size, \(N\) is the population size, and
\(f = (1 - n / N)\) (\(f\) is often called the finite population
correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used
to estimate means and the local neighborhood variance was used to
estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate
means and variances using restricted maximum likelihood. SRS, GRTS
sampling, and design-based inference were implemented using the spsurvey
\textbf{\textsf{R}} package (Dumelle et al., 2022). FPBK was implemented
using the sptotal \textbf{\textsf{R}} package (Higham et al., 2021a).

@@

This manuscript has a supplementary \textbf{\textsf{R}} package that
contains all of the data, code, and figures used in its creation. All
figures were created using the ggplot2 \textbf{\textsf{R}} package
(Wickham, 2016). The supplementary \textbf{\textsf{R}} package is hosted
on GitHub and archived on Zenodo. The Zenodo DOI is located here and in
the references.

@@

\leavevmode\hypertarget{ref-dumelle2022spsurvey}{}%
Dumelle, M., Kincaid, T.M., Olsen, A.R., Weber, M.H., 2022. Spsurvey:
Spatial sampling design and analysis.

>>>>

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less (be
more precise) than parameter estimates from samples lacking spatial
balance (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm to see widespread use was the generalized random tessellation
stratified (GRTS) algorithm (Stevens and Olsen, 2004). After the GRTS
algorithm was developed, several other spatially balanced sampling
algorithms emerged, including stratified sampling with compact
geographical strata (Walvoort et al., 2010), the local pivotal method
(Grafström et al., 2012; Grafström and Matei, 2018), spatially
correlated Poisson sampling (Grafström, 2012), balanced acceptance
sampling (Robertson et al., 2013), within-sample-distance sampling
(Benedetti and Piersimoni, 2017), and Halton iterative partitioning
sampling (Robertson et al., 2018). In this manuscript, we select
spatially balanced samples using the GRTS algorithm because it is
readily available in the spsurvey \textbf{\textsf{R}} package (Dumelle
et al., 2022b) and naturally accommodates finite and infinite sampling
frames, unequal inclusion probabilities, and replacement units.
Replacement units are additional population units that can be sampled
when a population unit originally selected can no longer be sampled. A
couple of reasons why an originally selected site can no longer be
sampled include its location being physically inaccessible or it is on
private land that the researcher does not have permission to access.

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate variances. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the response value of the \(i\)th
sampled population unit, \(\bar{z}\) is the mean of all \(z_i\), \(n\)
is the sample size, \(N\) is the population size, and
\(f = (1 - n / N)\) (\(f\) is often called the finite population
correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used
to estimate means and the local neighborhood variance was used to
estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate
means and variances using restricted maximum likelihood. SRS, GRTS
sampling, and design-based inference were implemented using the spsurvey
\textbf{\textsf{R}} package (Dumelle et al., 2022b). FPBK was
implemented using the sptotal \textbf{\textsf{R}} package (Higham et
al., 2021a).

@@

This manuscript has a supplementary \textbf{\textsf{R}} package that
contains all of the data, code, and figures used in its creation. All
figures were created using the ggplot2 \textbf{\textsf{R}} package
(Wickham, 2016). The supplementary \textbf{\textsf{R}} package is hosted
on GitHub and archived on Zenodo. The Zenodo DOI link is
https://doi.org/10.5281/zenodo.6633805 (Dumelle et al., 2022a).

@@

\leavevmode\hypertarget{ref-dumelle2022DvMsp}{}%
Dumelle, M., Higham, M., Hoef, J.M.V., Olsen, A.R., Madsen, L., 2022a.
USEPA/DvMsp: Publication release. Zenodo.
doi:\href{https://doi.org/10.5281/zenodo.6633805}{10.5281/zenodo.6633805}

\leavevmode\hypertarget{ref-dumelle2022spsurvey}{}%
Dumelle, M., Kincaid, T.M., Olsen, A.R., Weber, M.H., 2022b. Spsurvey:
Spatial sampling design and analysis.
